---
title: "The Story of America's Dam Infrastructure"
subtitle: "Exploring 92,428 Dams Across the United States"
format: 
  html:
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
execute:
  warning: false
  message: false
---

```{r setup}
#| include: false
library(readr)
library(dplyr)
library(ggplot2)
library(scales)
library(knitr)
library(janitor)

dat <- read_csv(here::here("data", "nation.csv"), skip = 1) |> 
  clean_names()

# Custom theme for plots
theme_dam <- theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 12, color = "gray50"),
    axis.text = element_text(size = 10),
    legend.position = "bottom"
  )
```

## 📊 Executive Summary

This analysis explores the **National Inventory of Dams (NID)** dataset, containing detailed information about **`r format(nrow(dat), big.mark = ",")`** dams across the United States. Our findings reveal fascinating patterns about America\'s water infrastructure, from the post-war construction boom to surprising geographic distributions.

::: {.callout-important}
## Key Findings at a Glance
- **Peak construction era**: 1960s saw `r format(dat |> filter(!is.na(year_completed)) |> mutate(decade = floor(year_completed / 10) * 10) |> count(decade) |> arrange(desc(n)) |> slice(1) |> pull(n), big.mark = ",")` dams built
- **Recreation dominates**: `r format(dat |> count(primary_purpose) |> arrange(desc(n)) |> slice(1) |> pull(n), big.mark = ",")` dams are primarily for recreation
- **Safety concerns**: `r format(sum(dat$hazard_potential_classification == "High", na.rm = TRUE), big.mark = ",")` dams classified as **high hazard**
- **Private ownership**: `r percent(sum(dat$primary_owner_type == "Private", na.rm = TRUE) / nrow(dat), accuracy = 1)` of dams are privately owned
:::

---

## 🏗️ The Great Dam Building Boom

### When Were America\'s Dams Built?

The story of American dam construction is essentially the story of **post-World War II infrastructure development**. The median dam was completed in **`r median(dat$year_completed, na.rm = TRUE)`**, highlighting the massive construction efforts of the mid-20th century.

```{r construction-timeline}
#| fig-cap: "Dam Construction by Decade: The 1960s Boom"
#| fig-height: 6

construction_data <- dat |>
  filter(!is.na(year_completed), year_completed > 1800) |>
  mutate(decade = floor(year_completed / 10) * 10) |>
  count(decade) |>
  mutate(
    peak_decade = decade == 1960,
    label = if_else(peak_decade, paste0(format(n, big.mark = ","), " dams"), "")
  )

ggplot(construction_data, aes(x = decade, y = n)) +
  geom_col(aes(fill = peak_decade), show.legend = FALSE) +
  geom_text(aes(label = label), vjust = -0.5, color = "darkred", fontface = "bold") +
  scale_fill_manual(values = c("FALSE" = "steelblue", "TRUE" = "darkred")) +
  scale_y_continuous(labels = comma_format()) +
  labs(
    title = "The 1960s: America\'s Dam Building Golden Age",
    subtitle = "Nearly 19,000 dams built in a single decade",
    x = "Decade",
    y = "Number of Dams Built",
    caption = "Source: National Inventory of Dams"
  ) +
  theme_dam
```

> **Fun Fact**: More dams were built in the **1960s alone** than in the previous 160 years combined!

---

## 🎯 What Are Dams Actually Used For?

Contrary to popular belief, **flood control** isn\'t the primary purpose of most American dams. Recreation is the most common purpose.

```{r purpose-analysis}
#| fig-cap: "Primary Purposes of American Dams"
#| fig-height: 8

purpose_data <- dat |>
  count(primary_purpose) |>
  arrange(desc(n)) |>
  slice(1:10) |>
  mutate(
    primary_purpose = case_when(
      primary_purpose == "Fire Protection, Stock, Or Small Fish Pond" ~ "Fire/Stock/Fish Pond",
      TRUE ~ primary_purpose
    ),
    primary_purpose = reorder(primary_purpose, n),
    percentage = n / sum(n) * 100
  )

ggplot(purpose_data, aes(x = primary_purpose, y = n)) +
  geom_col(fill = "forestgreen", alpha = 0.8) +
  geom_text(aes(label = paste0(format(n, big.mark = ","), "\n(", round(percentage, 1), "%)")), 
            hjust = -0.1, size = 3.5) +
  coord_flip() +
  scale_y_continuous(labels = comma_format(), expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Recreation Dominates American Dam Purposes",
    subtitle = "Top 10 primary purposes for the nation\'s 92,428 dams",
    x = "Primary Purpose",
    y = "Number of Dams",
    caption = "Source: National Inventory of Dams"
  ) +
  theme_dam
```

## The Surprising Recreation Story

**`r format(dat |> count(primary_purpose) |> arrange(desc(n)) |> slice(1) |> pull(n), big.mark = ",")`** dams (**`r round(100 * (dat |> count(primary_purpose) |> arrange(desc(n)) |> slice(1) |> pull(n)) / nrow(dat), 1)`%**) are primarily used for recreation. This includes:

- 🏊 Swimming and boating
- 🎣 Fishing ponds  
- 🏕️ Campground water features
- ⛳ Golf course ponds

---

## ⚠️ Safety First: High-Hazard Dams

The safety implications are sobering: **`r format(sum(dat$hazard_potential_classification == "High", na.rm = TRUE), big.mark = ",")`** dams are classified as **"High Hazard"**, meaning their failure would likely cause loss of life.

```{r hazard-map}
#| fig-cap: "High-Hazard Dams by State"
#| fig-height: 8

hazard_data <- dat |>
  filter(hazard_potential_classification == "High") |>
  count(state) |>
  arrange(desc(n)) |>
  slice(1:15) |>
  mutate(
    state = reorder(state, n),
    danger_level = case_when(
      n >= 1500 ~ "Extreme Risk",
      n >= 800 ~ "High Risk", 
      TRUE ~ "Moderate Risk"
    )
  )

ggplot(hazard_data, aes(x = state, y = n, fill = danger_level)) +
  geom_col() +
  geom_text(aes(label = format(n, big.mark = ",")), hjust = -0.1, size = 3) +
  coord_flip() +
  scale_fill_manual(
    values = c("Extreme Risk" = "darkred", "High Risk" = "orange", "Moderate Risk" = "gold"),
    name = "Risk Level"
  ) +
  scale_y_continuous(labels = comma_format(), expand = expansion(mult = c(0, 0.12))) +
  labs(
    title = "States with the Most High-Hazard Dams",
    subtitle = "These dams could cause loss of life if they fail",
    x = "State",
    y = "Number of High-Hazard Dams",
    caption = "Source: National Inventory of Dams"
  ) +
  theme_dam
```

::: {.callout-warning}
## Safety Spotlight
**North Carolina** leads with **`r format(dat |> filter(hazard_potential_classification == "High") |> count(state) |> arrange(desc(n)) |> slice(1) |> pull(n), big.mark = ",")`** high-hazard dams, followed by **Texas** with **`r format(dat |> filter(hazard_potential_classification == "High") |> count(state) |> arrange(desc(n)) |> slice(2) |> pull(n), big.mark = ",")`**.
:::

---

## 🏔️ Tallest Dams: Engineering Marvels

### America\'s Skyscrapers of Water

```{r tallest-dams}
#| fig-cap: "The 10 Tallest Dams in America"

tallest_data <- dat |>
  select(dam_name, state, dam_height_ft) |>
  filter(!is.na(dam_height_ft)) |>
  arrange(desc(dam_height_ft)) |>
  slice(1:10) |>
  mutate(
    dam_name = reorder(dam_name, dam_height_ft),
    dam_type = case_when(
      grepl("Slurry|Refuse|Tailings", dam_name, ignore.case = TRUE) ~ "Mining Waste",
      dam_name %in% c("Hoover Dam", "Glen Canyon Dam", "Oroville") ~ "Major Federal",
      TRUE ~ "Other"
    )
  )

ggplot(tallest_data, aes(x = dam_name, y = dam_height_ft, fill = dam_type)) +
  geom_col() +
  geom_text(aes(label = paste0(dam_height_ft, " ft")), hjust = -0.1, size = 3) +
  coord_flip() +
  scale_fill_manual(
    values = c("Major Federal" = "navy", "Mining Waste" = "brown", "Other" = "gray60"),
    name = "Dam Type"
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(
    title = "America\'s Tallest Dams: A Mix of Federal Projects and Mining Waste",
    subtitle = "Several mining waste dams rank among the nation\'s tallest structures",
    x = "Dam Name",
    y = "Height (Feet)",
    caption = "Source: National Inventory of Dams"
  ) +
  theme_dam
```

**Surprising finding**: Several of America\'s tallest "dams" are actually **mining waste containment structures** in Kentucky and West Virginia!

---

## 🏭 Who Owns America\'s Dams?

The ownership story might surprise you: **`r percent(sum(dat$primary_owner_type == "Private", na.rm = TRUE) / nrow(dat), accuracy = 1)`** of all dams are **privately owned**.

```{r ownership-pie}
#| fig-cap: "Dam Ownership Distribution"

ownership_data <- dat |>
  count(primary_owner_type) |>
  filter(!is.na(primary_owner_type), primary_owner_type != "Not Listed") |>
  arrange(desc(n)) |>
  mutate(
    percentage = n / sum(n) * 100,
    label = paste0(primary_owner_type, "\\n", format(n, big.mark = ","), " dams\\n(", round(percentage, 1), "%)")
  )

ggplot(ownership_data, aes(x = "", y = n, fill = primary_owner_type)) +
  geom_col(width = 1) +
  coord_polar("y", start = 0) +
  scale_fill_brewer(type = "qual", palette = "Set3") +
  theme_void() +
  theme(legend.position = "right") +
  labs(
    title = "Private Ownership Dominates American Dams",
    subtitle = "Nearly two-thirds of dams are privately owned",
    fill = "Owner Type",
    caption = "Source: National Inventory of Dams"
  )
```

---

## ⚡ Hydroelectric Hotspots

Where does America generate hydroelectric power? The answer might surprise you - it\'s not just the Pacific Northwest!

```{r hydro-map}
#| fig-cap: "Hydroelectric Dams by State"

hydro_data <- dat |>
  filter(primary_purpose == "Hydroelectric") |>
  count(state) |>
  arrange(desc(n)) |>
  slice(1:12) |>
  mutate(
    state = reorder(state, n),
    region = case_when(
      state %in% c("Maine", "New Hampshire", "Vermont", "Massachusetts") ~ "New England",
      state %in% c("New York") ~ "Mid-Atlantic",
      state %in% c("California", "Washington", "Oregon") ~ "West Coast",
      TRUE ~ "Other"
    )
  )

ggplot(hydro_data, aes(x = state, y = n, fill = region)) +
  geom_col() +
  geom_text(aes(label = n), hjust = -0.1, size = 3) +
  coord_flip() +
  scale_fill_manual(
    values = c("New England" = "darkgreen", "Mid-Atlantic" = "navy", 
               "West Coast" = "orange", "Other" = "gray60"),
    name = "Region"
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(
    title = "New York Leads in Hydroelectric Dams",
    subtitle = "New England states dominate small-scale hydroelectric generation",
    x = "State",
    y = "Number of Hydroelectric Dams",
    caption = "Source: National Inventory of Dams"
  ) +
  theme_dam
```

**Key insight**: **New England states** (Maine, New Hampshire, Vermont, Massachusetts) collectively have more hydroelectric dams than the traditionally hydro-focused Western states!

---

## 🌊 Average Dam Heights by State

Which states build the tallest dams on average?

```{r average-heights}
#| fig-cap: "Average Dam Height by State (States with 50+ dams)"

height_data <- dat |>
  filter(!is.na(dam_height_ft), dam_height_ft > 0) |>
  group_by(state) |>
  summarise(
    avg_height = round(mean(dam_height_ft), 1),
    count = n(),
    .groups = "drop"
  ) |>
  filter(count >= 50) |>
  arrange(desc(avg_height)) |>
  slice(1:15) |>
  mutate(
    state = reorder(state, avg_height),
    highlight = state %in% c("Washington", "West Virginia", "Colorado")
  )

ggplot(height_data, aes(x = state, y = avg_height, fill = highlight)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0(avg_height, " ft")), hjust = -0.1, size = 3) +
  coord_flip() +
  scale_fill_manual(values = c("FALSE" = "steelblue", "TRUE" = "darkred")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(
    title = "Washington State Builds the Tallest Dams on Average",
    subtitle = "Average height for states with 50+ dams in the database",
    x = "State",
    y = "Average Dam Height (Feet)",
    caption = "Source: National Inventory of Dams"
  ) +
  theme_dam
```

---

## 🔍 Strange and Fascinating Facts

### The Mystery of "Year 0" Dams

There are **`r sum(dat$year_completed == 0, na.rm = TRUE)`** dams in the dataset with a completion year of **0**. Let\'s investigate:

```{r year-zero}
year_zero_dams <- dat |>
  filter(year_completed == 0) |>
  select(dam_name, state, primary_purpose) |>
  slice(1:10)

if(nrow(year_zero_dams) > 0) {
  kable(year_zero_dams, caption = "Sample of Mysterious \'Year 0\' Dams")
} else {
  cat("No dams with year 0 found in this dataset.")
}
```

### Florida\'s Mining Surprise

Let\'s look at states with the most **tailings** (mining waste) dams:

```{r tailings-analysis}
#| fig-cap: "States with Most Mining Waste Dams"

tailings_data <- dat |>
  filter(primary_purpose == "Tailings") |>
  count(state) |>
  arrange(desc(n)) |>
  slice(1:10) |>
  mutate(state = reorder(state, n))

ggplot(tailings_data, aes(x = state, y = n)) +
  geom_col(fill = "brown", alpha = 0.8) +
  geom_text(aes(label = n), hjust = -0.1, size = 3) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(
    title = "Mining Waste Dams by State",
    subtitle = "Tailings dams for containing mining waste",
    x = "State",
    y = "Number of Tailings Dams",
    caption = "Source: National Inventory of Dams"
  ) +
  theme_dam
```

---

## 📈 Key Takeaways

::: {.callout-note collapse="false"}
## 🎯 What We Learned

1. **🏗️ Infrastructure Boom**: The 1960s were the golden age of dam construction
2. **🎣 Recreation Rules**: Most dams serve recreational purposes, not flood control
3. **⚠️ Safety Concerns**: Nearly 17,000 high-hazard dams need monitoring
4. **🏘️ Private Property**: Nearly 2/3 of dams are privately owned
5. **⚡ Hydro Surprise**: New England, not the West, leads in hydroelectric dam count
6. **🏔️ Height Champions**: Washington state builds the tallest dams on average
7. **⛏️ Mining Impact**: Mining waste dams are concentrated in specific states
:::

---

## 📊 Dataset Summary

**Data Source**: [National Inventory of Dams (NID)](https://nid.sec.usace.army.mil/#/downloads)  
**Maintained by**: U.S. Army Corps of Engineers  
**Total Records**: `r format(nrow(dat), big.mark = ",")`  
**Variables**: `r ncol(dat)`  
**Coverage**: All 50 US states plus territories  
**Analysis Date**: `r Sys.Date()`

---

## 📋 Data and Analysis

This analysis uses data from the **[National Inventory of Dams (NID)](https://nid.sec.usace.army.mil/#/downloads)**, maintained by the U.S. Army Corps of Engineers. The NID is a comprehensive database containing information on dams throughout the United States and its territories.

This analysis was prepared with the assistance of Anthropic Claude 4 Sonnet.

::: {.callout-note}
## About the National Inventory of Dams
The NID was established following the National Dam Safety Act of 1972 and serves as a key resource for dam safety, emergency preparedness, and water resource management. The database is regularly updated with information submitted by state dam safety agencies and federal agencies.
:::